---
title: "Program Analysis"
subtitle: "<a href='https://hub.crunchdao.com/competitions/broad-obesity-1' target='_blank'>Obesity Machine Learning Competition</a>"
author: "Julie Laffy<br><span class='affiliation'>Postdoctoral fellow of the Eric and Wendy Schmidt Center<br>Broad Institute of MIT and Harvard</span>"
date: "May 2026"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    theme: cosmo
    highlight: tango
    df_print: paged
---

<style>
body {
  font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto, "Helvetica Neue", Arial, sans-serif;
  font-size: 16px;
  line-height: 1.65;
  color: #2c3e50;
}
.main-container {
  max-width: 100%;
  padding-left: 30px;
  padding-right: 30px;
}
@media (min-width: 992px) {
  .main-container > .row > div.col-md-9,
  .main-container > .row > div.col-xs-12.col-sm-8.col-md-9 {
    max-width: 1100px;
  }
}
h1.title {
  font-size: 2.6em;
  font-weight: 700;
  color: #1f3a5f;
  margin-top: 0.5em;
  margin-bottom: 0.1em;
}
.subtitle, h3.subtitle {
  font-size: 1.55em;
  font-weight: 500;
  color: #2d5a8c;
  margin-top: 0;
  margin-bottom: 0.5em;
  padding-top: 0;
  border-bottom: none;
}
.subtitle a, h3.subtitle a {
  color: #2d5a8c;
  text-decoration: none;
  outline: none;
}
.subtitle a:focus:not(:focus-visible),
h3.subtitle a:focus:not(:focus-visible) {
  outline: none;
  box-shadow: none;
}
.subtitle a::after, h3.subtitle a::after {
  content: " \2197";
  font-size: 0.6em;
  font-weight: 700;
  vertical-align: 0.25em;
  margin-left: 3px;
  opacity: 0.75;
}
.subtitle a:hover::after, h3.subtitle a:hover::after {
  opacity: 1;
}
.subtitle a:hover, h3.subtitle a:hover {
  color: #1a4480;
}
h3.subtitle::after {
  content: '';
  display: block;
  width: 70px;
  border-bottom: 2px solid #cbd5e0;
  margin-top: 0.6em;
}
h1 {
  font-size: 1.9em;
}
h4.author {
  color: #4a5568;
  font-weight: 400;
  font-size: 1em;
  margin-top: 0.4em;
  margin-bottom: 0.2em;
}
h4.date {
  color: #718096;
  font-weight: 400;
  font-size: 0.9em;
  margin-top: 0;
  margin-bottom: 1.4em;
}
.affiliation {
  font-size: 0.85em;
  color: #718096;
}
h1, h2, h3 {
  color: #1f3a5f;
  font-weight: 600;
  padding-top: 0.25em;
}
h1 {
  border-bottom: 2px solid #e2e8f0;
  padding-bottom: 0.3em;
  margin-top: 1.6em;
}
h1:first-of-type {
  margin-top: 0.6em;
}
h2 {
  margin-top: 1.4em;
}
h3 {
  color: #2d5a8c;
}
a {
  color: #2b6cb0;
  text-decoration: none;
}
a:hover {
  color: #1a4480;
  text-decoration: underline;
}
code {
  background-color: #f4f6f8;
  color: #c7254e;
  padding: 1px 5px;
  border-radius: 3px;
  font-size: 0.9em;
}
pre {
  background-color: #f8f9fb;
  border: 1px solid #e4e8ed;
  border-radius: 5px;
  padding: 12px 14px;
  font-size: 0.88em;
  line-height: 1.5;
}
pre code {
  background-color: transparent;
  color: inherit;
  padding: 0;
  font-size: inherit;
}
blockquote {
  border-left: 4px solid #cbd5e0;
  background-color: #f7fafc;
  color: #4a5568;
  padding: 0.4em 1em;
  margin: 1em 0;
}
em {
  color: #4a5568;
}
.tocify {
  border-radius: 5px;
  border: 1px solid #e2e8f0;
  width: 290px;
  font-size: 0.92em;
}
@media (max-width: 991px) {
  .tocify { width: auto; }
}
@media (min-width: 992px) {
  div.col-md-3 { padding-left: 0; }
}
.tocify .list-group-item {
  border-color: #edf2f7;
  white-space: normal;
  word-wrap: break-word;
  line-height: 1.35;
  padding-top: 7px;
  padding-bottom: 7px;
}
.tocify .tocify-subheader > li > a {
  white-space: normal;
  padding-top: 5px;
  padding-bottom: 5px;
}
.tocify .tocify-header .tocify-item.active,
.tocify .tocify-subheader .tocify-item.active {
  background-color: #2b6cb0;
  color: white;
  border-left-color: #1a4480;
}
.tocify .list-group-item:hover {
  background-color: #f7fafc;
}
hr {
  border-top: 1px solid #e2e8f0;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE, message = FALSE, warning = FALSE)
```

# Overview

Program analysis methodology used in the Obesity ML Competition.

- **Parts 1–3** — cell-program classification (pre-adipocyte / adipocyte /
  lipogenic / other) and proportions. Methodology used for
  [Crunch 1](https://hub.crunchdao.com/competitions/broad-obesity-1) and
  [Crunch 2](https://hub.crunchdao.com/competitions/broad-obesity-2).
- **Part 4** — thermogenic signature scoring. Methodology used for
  [Crunch 3](https://hub.crunchdao.com/competitions/broad-obesity-3).

# Setup

```{r}
# Install and load required packages for this analysis

# Set CRAN mirror (needed for non-interactive/vanilla mode)
options(repos = c(CRAN = "https://cloud.r-project.org"))

# CRAN packages
cran_packages <- c("Matrix", "dplyr", "tidyr", "reshape2")

# Install missing CRAN packages
install.packages(setdiff(cran_packages, rownames(installed.packages())))

# BiocManager for Bioconductor packages (scalop depends on these)
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

# Install scalop
if (!"scalop" %in% rownames(installed.packages())) {
  BiocManager::install("jlaffy/scalop")
}

# Load all
invisible(lapply(c(cran_packages, "scalop"), library, character.only = TRUE))
```

# Part 1: All Cells

## 1.1 Preprocess

```{r}
library(Matrix)
library(scalop)

full_matrix = "data/m.allgenes.logcpm.rds"
proc_matrix = "data/m.logcpm.rds" 
gene_cutoff = 3

# read in expression matrix
# already in form log2(CPM/10 + 1)
m <- readRDS(full_matrix)
m <- as.matrix(m)

# calculate average gene expression
avg <- scalop::aggr_gene_expr(m, isBulk=F)

# filter genes that do not pass minimum gene expression cutoff
m2 <- m[avg >= gene_cutoff, ]

# save filtered matrix
saveRDS(m2, proc_matrix)
```

## 1.2 Score Adipo/Preadipo

```{r}
proc_matrix = "data/m.logcpm.rds" 
gene_sigs = "sigs/adipo_preadipo_sigs.rds"
frac_genes_retained = 0.5
n_genes_retained = 10
score_results = "data/permuted_scores.adipo_preadipo_sigs.rds"

# Read in processed expression matrix
m = readRDS(proc_matrix)

# Read in gene signatures to score cells by
# there are 2 adipocyte signatures: "adipomap.Ad" and "yi.AD"
# and 2 pre-adipocyte (progenitor) signatures: "adipomap.ASPC" and "yi.ASPC"
sigs = readRDS(gene_sigs)

# Filter out signatures whose genes are poorly represented in {proc_matrix}
# A signature is kept if at least 10 genes and 50% of the signature exist in {proc_matrix}
# {frac_genes_retained} and {n_genes_retained}
sigs_filt = sigs[sapply(sigs, function(x) sum(x %in% rownames(m)) >= n_genes_retained)]

# Calculate single-cell expression scores for each signature and FDRs
invisible(capture.output(
  scores_permuted <- scalop::permuteSigScores(m, sigs_filt, conserved.genes=frac_genes_retained)
))

# Save results
saveRDS(scores_permuted, score_results)
```

## 1.3 Classify Adipo/Preadipo/Other

```{r}
score_results = "data/permuted_scores.adipo_preadipo_sigs.rds"
fdr_cutoff = 0.01 
diff_cutoff = 0.3

# Load scores
scores = readRDS(score_results)

# Prepare data
scores_wide = scores %>%
  dplyr::select(id, sig, score, fdr) %>%
  tidyr::pivot_wider(names_from = sig, values_from = c(score, fdr)) %>%
  dplyr::mutate(
    max_adipo = pmax(score_adipomap.Ad, score_yi.AD),
    max_preadipo = pmax(score_adipomap.ASPC, score_yi.ASPC),
    score_diff = max_adipo - max_preadipo,
    max_overall = pmax(score_adipomap.Ad,
                       score_yi.AD, 
                       score_adipomap.ASPC, 
                       score_yi.ASPC)
  )

# Classify cells
# Adipo logic: 
# - score_diff > {diff_cutoff} (adipo score higher than preadipo score)
# - both ASPC (preadipo) signatures must be non-significant (fdr >= {fdr_cutoff})
# - at least one adipocyte signature must be significant ({fdr_cutoff})
# Preadipo logic: 
# - score_diff < -{diff_cutoff} (preadipo score higher than adipo score)
# - both adipo signatures must be non-significant (fdr >= {fdr_cutoff})
# - at least one ASPC (preadipocyte) signature must be significant ({fdr_cutoff})
# Other: everything else (ambiguous, dual-identity, neither) 

scores_wide = scores_wide %>%
  dplyr::mutate(
    compartment = dplyr::case_when(
      score_diff > diff_cutoff & 
        (fdr_adipomap.ASPC >= fdr_cutoff & fdr_yi.ASPC >= fdr_cutoff) &
        (fdr_adipomap.Ad < fdr_cutoff | fdr_yi.AD < fdr_cutoff) ~ "Adipo",
      score_diff < -1 * diff_cutoff & 
        (fdr_adipomap.Ad >= fdr_cutoff & fdr_yi.AD >= fdr_cutoff) &
        (fdr_adipomap.ASPC < fdr_cutoff | fdr_yi.ASPC < fdr_cutoff) ~ "Preadipo",
      TRUE ~ "Other"
    )
  )				

# Extract cell IDs for each compartment
adipo_cells = scores_wide$id[scores_wide$compartment == "Adipo"]
preadipo_cells = scores_wide$id[scores_wide$compartment == "Preadipo"]
other_cells = scores_wide$id[scores_wide$compartment == "Other"]

# Save cell ID lists
saveRDS(adipo_cells, 'data/cells.adipo.rds')
saveRDS(preadipo_cells, 'data/cells.preadipo.rds')
saveRDS(other_cells, 'data/cells.other.rds')
```

# Part 2: Adipo Cells

## 2.1 Preprocess

```{r}
adipo_cells = "data/cells.adipo.rds"
full_matrix = "data/m.allgenes.logcpm.rds"
proc_matrix = "data/m.adipo.logcpm.rds"
gene_cutoff = 3

# read expression matrix and cells classified as Adipo
cells = readRDS(adipo_cells)
m = readRDS(full_matrix)
m = as.matrix(m)

# filter to keep only Adipo cells 
m_adipo = m[, cells]

# calculate average gene expression
avg = scalop::aggr_gene_expr(m_adipo, isBulk=F)

# filter genes that do not pass minimum gene expression cutoff
m2 = m_adipo[avg >= gene_cutoff, ]

# save filtered matrix
saveRDS(m2, proc_matrix)
```

## 2.2 Score Lipogenic

```{r}
proc_matrix = "data/m.adipo.logcpm.rds" 
gene_sigs = "sigs/lipogenic_sigs.rds"
frac_genes_retained = 0.5
n_genes_retained = 10
score_results = "data/permuted_scores.lipogenic_sigs.rds"

# Read in processed expression matrix
m = readRDS(proc_matrix)

# Read in gene signatures to score cells by
# These are a collection of lipogenic signatures
sigs = readRDS(gene_sigs)

# Filter out signatures whose genes are poorly represented in {proc_matrix}
# A signature is kept if at least 10 genes and 50% of the signature exist in {proc_matrix}
# {frac_genes_retained} and {n_genes_retained}
sigs_filt = sigs[sapply(sigs, function(x) sum(x %in% rownames(m)) >= n_genes_retained)]

# Calculate single-cell expression scores for each signature and FDRs
invisible(capture.output(
  scores_permuted <- scalop::permuteSigScores(m, sigs_filt, conserved.genes=frac_genes_retained)
))

# Save results
saveRDS(scores_permuted, score_results)
```

## 2.3 Classify Lipogenic

```{r}
score_results = "data/permuted_scores.lipogenic_sigs.rds"
fdr_cutoff = 0.01
score_cutoff = 0.3
n_sigs = 2

# Load scores
scores = readRDS(score_results)

# Prepare data
scores_wide = scores %>%
  dplyr::select(id, sig, score, fdr) %>%
  tidyr::pivot_wider(names_from = sig, values_from = c(score, fdr))

# Count how many signatures pass criteria per cell
n_pass = rowSums(sapply(names(scores_wide)[grepl("^score_", names(scores_wide))], function(score_col) {
  fdr_col = sub("^score_", "fdr_", score_col)
  scores_wide[[score_col]] > score_cutoff & scores_wide[[fdr_col]] < fdr_cutoff
}))

# Filter for cells that have at least {n_sigs} lipogenic signatures passing cutoffs
lipogenic_cells = scores_wide$id[n_pass >= n_sigs]

# Save cell ID list
saveRDS(lipogenic_cells, "data/cells.lipogenic.rds")
```

# Part 3: Proportions

```{r}
# get cell assignment file paths
cell_files = list.files("data", full.names = T, pattern = "^cells")

# read in cell assignments
cell_list = sapply(cell_files, readRDS, simplify=F)

# fix names and reorder groups
names(cell_list) = scalop::substri(basename(names(cell_list)), pos=2)
ord = c('adipo', 'preadipo', 'other', 'lipogenic')
cell_list = cell_list[ord]

# get cell proportions
n_cells = lengths(cell_list)
frac_cells = n_cells / sum(n_cells[-4])
frac_cells
```

*Expected output for sample data:*

```
##     adipo  preadipo     other lipogenic
##     0.291     0.384     0.325     0.084
```

# Part 4: Thermogenic Scoring

The scoring pipeline below is what produced the reference
`*_ThermoScores_cell.csv` and `*_ThermoScores_perturbation.csv` files released
to participants in
[Crunch 3](https://hub.crunchdao.com/competitions/broad-obesity-3). Twelve
signatures are scored, taken from `thermogenic_signatures.csv`:

```
nonucp1.all
REACTOME_AMPK_inhibits_chREBP_R-HSA-163680
REACTOME_Integration_of_energy_metabolism_R-HSA-163685
REACTOME_Mitochondrial_biogenesis_R-HSA-1592230
emont.hAd6
lit.thermogenic
C2.WP_THERMOGENESIS
C5.GOBP_ADAPTIVE_THERMOGENESIS
C5.GOBP_BROWN_FAT_CELL_DIFFERENTIATION
C5.GOBP_NEGATIVE_REGULATION_OF_COLD_INDUCED_THERMOGENESIS
C5.GOBP_POSITIVE_REGULATION_OF_COLD_INDUCED_THERMOGENESIS
yi.Thermogenesis_Village
```

## 4.1 Score Thermogenic Signatures

```{r}
proc_matrix = "data/m.adipo.logcpm.rds"
gene_sigs = "sigs/thermo_sigs.rds"
n_genes_retained = 2
score_results = "data/permuted_scores.thermo_sigs.rds"

# Read in processed adipocyte expression matrix
m = readRDS(proc_matrix)

# Read in the 12 thermogenic gene signatures
sigs = readRDS(gene_sigs)

# Trim each signature to genes present in the matrix
sigs_present = sapply(sigs, function(x) x[x %in% rownames(m)], simplify = FALSE)

# Keep only sigs with at least {n_genes_retained} of their genes in the matrix
sigs_to_score = sigs_present[lengths(sigs_present) >= n_genes_retained]

# Score each signature against each cell (raw score + permutation-based FDR)
scores = permuteSigScores(m, sigs_to_score, conserved.genes = 0)

# Annotate each row with how many genes the signature contributed
scores$n_genes_used = lengths(sigs_to_score)[as.character(scores$sig)]

# Save long per-(cell, sig) table: id, sig, score, fdr, n_genes_used
saveRDS(scores, score_results)
```

## 4.2 Z-score and Per-Perturbation Aggregation

Z-score each signature across all adipocyte cells, aggregate to perturbation
level (mean z per gene × sig), and rank perturbations by
`agg_top3_z` (mean of the top-3 signature z-scores). Outputs in the format of
the [released reference files](https://hub.crunchdao.com/competitions/broad-obesity-3)
for Crunch 3: `ThermoScores_cell.csv` (one row per cell) and
`ThermoScores_perturbation.csv` (one row per perturbation, with `agg_top3_z`
and `rank`).

```{r}
score_results = "data/permuted_scores.thermo_sigs.rds"
cell_out = "data/ThermoScores_cell.csv"
perturbation_out = "data/ThermoScores_perturbation.csv"

# Load per-(cell, sig) scores from script 08
scores = readRDS(score_results)

# 1. Z-score each signature across all adipocyte cells
# z = (raw_score - mean_across_cells) / sd_across_cells
scores = scores %>%
  dplyr::group_by(sig) %>%
  dplyr::mutate(z_score = (score - mean(score)) / stats::sd(score)) %>%
  dplyr::ungroup()

# 2. Derive perturbation (gene) from cell id prefix.
#    TF150 convention only — otherwise use AnnData `.obs['gene']` / cell metadata.
scores$gene = sub("_.*", "", as.character(scores$id))

# 3. Per-cell wide table: raw score columns + z_-prefixed z-score columns
cell_wide = scores %>%
  dplyr::select(cell_id = id, gene, sig, score, z_score) %>%
  tidyr::pivot_wider(names_from = sig, values_from = c(score, z_score),
                     names_glue = "{ifelse(.value == 'score', '', 'z_')}{sig}")

write.csv(cell_wide, cell_out, row.names = FALSE)

# 4. Per-perturbation aggregation:
#    - mean z-score per (gene, sig) across all cells of that perturbation
#    - agg_top3_z = mean of the top-3 sig z-scores
#    - rank perturbations from highest to lowest agg_top3_z
perturb_long = scores %>%
  dplyr::group_by(gene, sig) %>%
  dplyr::summarise(mean_z = mean(z_score), .groups = "drop")

perturb_wide = perturb_long %>%
  tidyr::pivot_wider(names_from = sig, values_from = mean_z)

agg_top3 = function(row) mean(sort(unlist(row), decreasing = TRUE)[1:3])

sig_cols = setdiff(names(perturb_wide), "gene")
perturb_wide$agg_top3_z = apply(perturb_wide[, sig_cols], 1, agg_top3)
perturb_wide = perturb_wide %>% dplyr::arrange(dplyr::desc(agg_top3_z))
perturb_wide$rank = seq_len(nrow(perturb_wide))

write.csv(perturb_wide, perturbation_out, row.names = FALSE)
```

*Below: results from running the pipeline on the **TF150** dataset.*

**`ThermoScores_cell.csv`** — first 3 of 30,975 rows, showing 2 of the
12 raw + 12 z-scored signature columns:

```
                    cell_id  gene  emont.hAd6  z_emont.hAd6
TRIM5_P1_AAACCATTCGCGACAT-1 TRIM5      0.6472        1.9737
CREB1_P1_AAACCCTGTAGTAGGC-1 CREB1      0.1894        0.5775
TRRAP_P1_AAACGAATCAATTGGT-1 TRRAP      0.0647        0.1972
```

**`ThermoScores_perturbation.csv`** — top 8 of 158 perturbations ranked by
`agg_top3_z`, showing the first 2 of the 12 signature columns:

```
    gene  emont.hAd6  lit.thermogenic  agg_top3_z  rank
RNASEH2C      0.5796           0.1700      0.5484     1
 FAM136A      0.1190           0.4887      0.4885     2
 TMEM107     -0.2030           0.3624      0.4199     3
   EP400      0.2438           0.3996      0.3956     4
   TRIM5      0.1003           0.1147      0.3177     5
   CEBPG      0.0505           0.0977      0.2926     6
   MEF2A     -0.0477           0.2802      0.2907     7
    RBAK      0.1003           0.3238      0.2900     8
```

